---
output: html_document
editor_options: 
  chunk_output_type: console
---
# Rural schools have declining enrollment and small class sizes

```{r libraries}
#| echo: false
#| warning: false
#| output: false

source(here::here("r", "libraries.r"))

```


```{r locations}
#| echo: false
#| warning: false
#| output: false

dpad <- r"(E:\data\cornell_pad\)"
dschools <- path(dpad, "schools")
dnysed <- r"(E:\data\nyschools\)"

```


```{r getdata}
#| echo: false
#| warning: false
#| output: false

df <- readRDS(here::here("data", "enroll_marked.rds"))

```


```{r data_prep}
#| echo: false
#| warning: false
#| output: false

df2 <-df |> 
  filter(!is.na(locale)) |> # remove nonmatches
  arrange(year, districtid) |> 
  group_by(year, districtid) |> 
  mutate(count=n()) |> 
  ungroup()
# count(df2, count) # good - no dups

df3 <- df2 |> 
  mutate(locale=as.integer(locale),
        locgrp=case_when(locale %in% 11:13 ~ "city",
                 locale %in% 21:23 ~ "suburban",
                 locale %in% 31:33 ~ "town",
                 locale %in% 41:43 ~ "rural",
                 TRUE ~ "error"))
# count(df3, locgrp, locale, localef)
count(df3 |> filter(year==max(year)), locgrp, locale, localef)

```


## Rural enrollment has been declining rapidly

Sources: NYSED enrollment data obtained via Cornell PAD. Rural/suburban/town designations from NCES. (Details to come.)

```{r}
#| echo: false
#| warning: false
#| message: false
#| notes: false
#| output: true

options(dplyr.summarise.inform = FALSE)
p <- df3 |>
  filter(locgrp != "city") |> 
  group_by(year, locgrp) |>
  summarise(value=median(total)) |> 
  ggplot(aes(year, value / value[year==1995], colour=locgrp)) +
  geom_line() +
  geom_point() +
  geom_hline(yintercept = 1) +
  scale_y_continuous(name="% of 1995 enrollment", labels=scales::percent_format(accuracy = 1)) +
  scale_x_continuous(name="Year in which school year begins", breaks=seq(1990, 2025, 5)) +
  ggtitle("Median enrollment in NY rural, suburban and town school districts",
          subtitle="Enrollment indexed to 1995=100") +
  theme_bw() +
  legend_notitle
ggsave(here::here("results", "enroll_indexed.png"), p, width=8, height=6)
options(dplyr.summarise.inform = TRUE)

```


```{r}
count(df3, locgrp)
tmp <- df3 |>
  filter(locgrp == "rural") |> 
  select(year, cnty, districtid, leaid, legalname, year, localef, locgrp, total) |> 
  filter(year %in% c(1995, 2021)) |> 
  pivot_wider(names_from = year, values_from = total) |> 
  mutate(change=`2021` - `1995`)

tmp |> 
  mutate(type=case_when(is.na(change) ~ "NA",
                        change < 0 ~ "decline",
                        change > 0 ~ "increase",
                        change == 0 ~ "zero",
                        TRUE ~ "ERROR"
                        )) |> 
  count(type)

```


## Rural school districts have smaller enrollment than other districts

Median enrollment is now less than 750 students - slightly more than half of town school districts, and about a quarter of suburban school districts.


```{r}
#| echo: false
#| warning: false
#| message: false
#| notes: false
#| output: true

options(dplyr.summarise.inform = FALSE)
df3 |>
  filter(locgrp != "city") |> 
  group_by(year, locgrp) |>
  summarise(value=median(total)) |> 
  ggplot(aes(year, value, colour=locgrp)) +
  geom_line() +
  geom_point() +
  scale_y_continuous(name="Fall enrollment", labels=scales::number_format(accuracy = 1), 
                     limits=c(0, NA), breaks=seq(0, 5000, 250)) +
  scale_x_continuous(name="School year beginning in...", breaks=seq(1990, 2025, 5)) +
  ggtitle("Median enrollment in NY rural, suburban and town school districts") +
  theme_bw()
options(dplyr.summarise.inform = TRUE)

```

## District-wide enrollment in 12th grade

Median 12th grade enrollment in NY rural school districts is about 50 students - about half of town districts and a fifth of suburban districts. It is down about a third from its level in 2007.

This makes it more difficult to run courses that might be of interest to a relatively small number of students, such as AP courses, less-popular foreign languages, and some vocational courses.


```{r}
#| echo: false
#| warning: false
#| message: false
#| notes: false
#| output: true

options(dplyr.summarise.inform = FALSE)
df3 |>
  filter(locgrp != "city") |> 
  group_by(year, locgrp) |>
  summarise(value=median(g12)) |> 
  ggplot(aes(year, value, colour=locgrp)) +
  geom_line() +
  geom_point() +
  scale_y_continuous(name="Fall enrollment", labels=scales::number_format(accuracy = 1), 
                     limits=c(0, NA), breaks=seq(0, 300, 25)) +
  scale_x_continuous(name="School year beginning in...", breaks=seq(1990, 2025, 5)) +
  ggtitle("Median 12th grade enrollment in NY rural, suburban and town school districts") +
  theme_bw()
options(dplyr.summarise.inform = TRUE)
```


