---
output: html_document
editor_options: 
  chunk_output_type: console
---

# Cornell data

This page documents data sources obtained from Cornell's Program on Applied Demographics (PAD) program.


New York State Center for Rural Schools https://www.nyruralschools.org/ Data https://www.nyeducationdata.org/

I read all data as character to make it easier to look for bad data before converting to numeric.


```{r }
#| label: setup
#| eval: true
#| include: false
#| echo: false

source(here::here("r", "libraries.r"))
dpad <- r"(E:\data\cornell_pad\)"
dschools <- path(dpad, "schools")

```

## Download files

https://pad.human.cornell.edu/schools/datadownload.cfm

Download the following files:



```{r}
#| label: files-and-locations
#| eval: true
#| include: true
#| echo: false
#| df-print: tibble


# last download: 6/12/2023  7/20/2022
# on left is the name Cornell associates with a file, on right is its url

fnames <- read_csv("fname, url
basicinfo.csv,            https://pad.human.cornell.edu/schools/BasicInfo_csv.cfm
demographics_all.csv,     https://pad.human.cornell.edu/schools/demogr_csv.cfm
enrollments_all.csv,      https://pad.human.cornell.edu/schools/enr_csv.cfm
graduation_all.csv,       https://pad.human.cornell.edu/schools/grad_csv.cfm
regents_all.csv,          https://pad.human.cornell.edu/schools/regents_csv.cfm
apm_all.csv,              https://pad.human.cornell.edu/schools/apm_csv.cfm
faru_all.csv,             https://pad.human.cornell.edu/schools/FARU_csv.cfm
subgroups.csv,            https://pad.human.cornell.edu/schools/download/Subgroups.csv
elamath_all.csv,          https://pad.human.cornell.edu/schools/ELAMATH_csv.cfm
")

options(width=120)
print(fnames |> arrange(fname))
options(width=80) # default

```


```{r}
#| label: ONETIME-downloads
#| eval: false
#| include: false
#| echo: false

# download files ----------------------------------------------------------

# https://pad.human.cornell.edu/schools/datadownload.cfm 

# Please note that Year in the data files refers to the beginning year of the
# school year. E.g. 2010 refers to 2010/2011 school year.


# note that:
#   I put ELAMATH last because it might timeout -- change the timeout option as shown below
#   subgroups.csv does not have a .cfm link
#   some of the files can take a long time, especially graduation_all.csv and regents_all.csv

# https://pad.human.cornell.edu/schools/BasicInfo_csv.cfm

# download some or all files
tout <- getOption("timeout")
options(timeout=180)
for(fnum in 1:nrow(fnames)){
# for(fnum in 8){
  fname <- fnames$fname[fnum] |> str_to_lower()
  url <- fnames$url[fnum]
  print(fname)
  download.file(url, path(dschools, fname), mode="wb")
}
options(timeout=tout)


```


## basic info

This file has district id, name, BOCES code, grades served, and needs index for 2016-17.

```{r}
#| label: basic-info
#| eval: false
#| include: false
#| echo: false

basic1 <- vroom(path(dschools, "basicinfo.csv"),
               col_types = cols(.default = col_character()))
glimpse(basic1)
count(basic1, YEAR) # all data are from 2016-17
skim(basic1)
# GEOID and DSB have some missing values

basic2 <- basic1 |>
  lcnames() |>
  rename(distid=districtid,
         dname=district_name, 
         grades=grade_range, 
         ineeds=needs_index, 
         boces=boces_cd) |>
  mutate(ineeds=as.integer(ineeds)) |>
  select(distid, dsb, geoid, dname, boces, grades, ineeds)
glimpse(basic2)
count(basic2, ineeds) # 1-6


# do we need dsb AND distid?
basic2 |> filter(distid != dsb) # none
basic2 |> filter(is.na(distid)) # none
basic2 |> filter(is.na(dsb)) # about 8
# no, we can drop dsb

basic <- basic2 |>
  # make nicer district names
  mutate(dname=str_to_title(dname),
         dname=str_replace(dname,  " Csd", " CSD"),
         dname=str_replace(dname,  " Sd", " SD"),
         dname=str_replace(dname,  "Ufsd", "UFSD"), # note no space so replace anywhere
         dname=str_replace(dname,  "Nyc", "NYC")
  ) |> 
  select(-dsb)
unique(basic$dname) |> sort()
glimpse(basic)
saveRDS(basic, here::here("data", "basic.rds"))

summary(basic)
skim(basic)

```

```{r}
#| label: basic-show
#| eval: true
#| include: true

# note that the following option will use more of the page:
#   #| column: page

basic <- readRDS(here::here("data", "basic.rds"))
glimpse(basic)

options(width = 120) 
print(skim(basic))
options(width=80) # default


```

## demographics

Cornell's source:

-   Cornell (Jan Vink) obtains the data from [nysed](https://www.p12.nysed.gov/irs/statistics/enroll-n-staff/ArchiveEnrollmentData.html).

-   Note that some of the web data are slightly updated vs Jan's data

```{r}
#| label: demographics
#| eval: false
#| include: false
#| echo: false

demographics1 <- vroom(path(dschools, "Demographics_all.csv"),
                 col_types = cols(.default = col_character()))
glimpse(demographics1)

demographics2 <- demographics1 |>
  lcnames() |> 
  rename(distid=district) |>
  mutate(year=as.integer(year),
         across(-c(distid, year), as.numeric))
glimpse(demographics2)
print(skim(demographics2))

demographics2 |> 
  filter(year==2019, distid=="010100")

saveRDS(demographics2, here::here("data", "demographics_wide.rds"))

# subgroups, labels; vname
# 01  All Students; ? not included in this data, but in enrollment instead
# 02	Female; female
# 03	Male; male

# 04	American Indian/Alaska Native; amind
# 05	Black; black
# 06	Hispanic; hisp
# 07	Asian/Pacific Islander; asian
# 08	White; white
# 09	Multiracial; multi

# 10	General Education Students; ?? by subtraction?
# 11	Students with Disabilities; swd

# 12	Not English Language Learner; ?? by subtraction ?
# 13	English Language Learner; ell

# 15	Economically Disadvantaged; ecdis
# 16	Not Economically Disadvantaged; ???? - by subtraction?

# not found: lep, freelunch, reducedlunch
# lep: Limited English Proficient subgroup 12

# saveRDS(demographics, here::here("data", "demographics.rds"))
```


```{r}
#| label: demographics-show
#| eval: true
#| include: true


demo <- readRDS(here::here("data", "demographics_wide.rds"))
glimpse(demo)

options(width = 120) 
print(skim(demo))
options(width=80) # default

```


## enrollment

Enrollment by district by grade by year, in wide format:

-   Cornell (Jan Vink) obtains the data from two nysed web pages: [this](https://www.p12.nysed.gov/irs/statistics/enroll-n-staff/ArchiveEnrollmentData.html), and [this](https://www.p12.nysed.gov/irs/statistics/enroll-n-staff/home.html) -- the latter is the main page.

-   Note that some of the web data are slightly updated vs Jan's data

-   Cornell only has some subsets of the data, not the crosses (e.g., not white-female)


```{r}
#| label: enrollment
#| eval: false
#| include: false
#| echo: false

enroll1 <- vroom(path(dschools, "enrollments_all.csv"),
              col_types = cols(.default = col_character()))
glimpse(enroll1)

enroll2 <- enroll1 |>
  lcnames() |> 
  rename(distid=distrid) |>
  mutate(year=as.integer(year),
         across(-c(distid, year), as.numeric))
glimpse(enroll2)

#.. enrollment wide ----
saveRDS(enroll2, here::here("data", "enrollwide.rds"))

#.. enrollment long ----
# enroll <- enroll2 |>
#   pivot_longer(cols=-c(distid, year))
# ht(enroll)
# saveRDS(enroll, here::here("data", "enroll.rds"))


```

```{r}
#| label: enrollment-show
#| eval: true
#| include: true

enrollwide <- readRDS(here::here("data", "enrollwide.rds"))
glimpse(enrollwide)

options(width = 120) 
print(skim(enrollwide))
options(width=80) # default

```


## ELA and Math

CAUTIONS:

-   The variable that NYSED calls `TOTAL_SCORE` has many negative values.

-   It looks like missing values for levels are coded as -1.


I convert:

-   Negative levels, total score, and average scores to missing


```{r}

elamath1 <- vroom(path(dschools, "elamath_all.csv"),
                  col_types = cols(.default = col_character()))
glimpse(elamath1)
count(elamath1, SUBJECT)
count(elamath1, SUBGROUP)
count(elamath1, GRADE)
count(elamath1, YEAR)

elamath2 <- elamath1 |>
  lcnames() |> 
  rename(distid=beds_cd, 
         numtested=num_tested, 
         totscore=total_score) |>
  mutate(distid=str_pad(distid, width=6, side="left", pad="0"),
         across(c(year, subgroup, grade, numtested), as.integer),
         across(c(starts_with("level"), totscore, avgscore), as.numeric)) |>
  select(distid, year, subject, grade, subgroup, everything())
summary(elamath2) # looks like -1 should be set to NA
quantile(elamath2$totscore)
elamath2 |> filter(totscore < 0) # totscore looks suspicious

elamath3 <- elamath2 |>
  mutate(across(c(starts_with("level"), avgscore), ~ ifelse(.x==-1, NA_real_, .x)),
         level12=naz(level1) + naz(level2),
         level34=naz(level3) + naz(level4),
         totscore=ifelse(totscore < 0, NA_real_, totscore),
         avgscore=ifelse(avgscore < 0, NA_real_, avgscore))
summary(elamath3)

# elamath <- elamath3 |>
#   pivot_longer(cols=c(numtested,
#                       starts_with("level"),
#                       totscore, avgscore))

saveRDS(elamath3, here::here("data", "elamath_wide.rds"))


```



## faru (finance) data

FARU is the Fiscal Analysis and Research Unit in NYSED.

```{r}
#| label: finance
#| eval: false
#| include: false
#| echo: false

finance1 <- vroom(path(dschools, "faru_all.csv"),
                       col_types = cols(.default = col_character()))
glimpse(finance1)

finance2 <- finance1 |>
  lcnames() |> 
  rename(distid=district) |>
  mutate(year=as.integer(year),
         across(-c(distid, year), as.numeric))
glimpse(finance2)
names(finance2)

#.... finance wide Cornell ----
saveRDS(finance2, here::here("data", "finwide.rds"))

```


```{r}
#| label: finance-show
#| eval: true
#| include: true

finwide <- readRDS(here::here("data", "finwide.rds"))
glimpse(finwide)

options(width = 120) 
print(skim(finwide))
options(width=80) # default

```

## graduation rates

http://www.nysed.gov/news/2022/state-education-department-releases-2017-cohort-high-school-graduation-rates 2017 cohort, those students who first entered 9th grade in New York's public schools in 2017 this means: cohortye is the year a student first entered 9th grade in NY public schools


## subgroups

```{r}
#| label: subgroups
#| eval: false
#| include: false
#| echo: false

subgroups <- vroom(path(dschools, "subgroups.csv")) |>
  lcnames() |> 
  mutate(subgroup=as.integer(subgroup))
subgroups
saveRDS(subgroups, here::here("data", "subgroups.rds"))

```

```{r}
#| label: subgroups-show
#| eval: true
#| include: true

subgroups <- readRDS(here::here("data", "subgroups.rds"))
glimpse(subgroups)

options(width = 120) 
print(skim(subgroups))
options(width=80) # default

count(subgroups, subgroup, description)

```

